Monitoring Time Series With Missing Values:\\a Deep Bayesian Approach

### 2022-02-15 paper.tex
intro in progress

Evolving state of OT environments  is monitored through collection and
streaming of multivariate time series. Advances in time series forecasting due
to adoption of multilayer recurrent neural network architectures make it
possible to forecast in high-dimensional time series, and identify and
classify novelties (anomalies) early, based on subtle changes in the trends.
However, mainstream approaches to multi-variate time series predictions do not
handle well cases when the ongoing forecast must include uncertainty and shape
(possibly multimodal) of future distributions of observations. Stochastic
latent variable variants of high-dimensional time series models where
proposed, but so far have had to rely on sampling to account for uncertainty,
limiting the performance of data handling. In addition, main-stream methods of
multivariate time series forecasting are not well-suited for handling missing
data, while transient unavailability of some of the data sources (e.g. when a
sensor stops working, or a transport channel malfunctions) is  a common case
with monitoring of complex systems.  Combining state-of-the-art methods of
forecasting in high-dimensional time series (based on neural-network
architectures) with full Bayesian handling of uncertainty (e.g. with help of
deep probabilistic programming systems such as Pyro, Tensorflow Probability,
and others) will allow to achieve robust forecasting (a precondition for
anomaly detection) in high-dimensional time series with missing data.
Representing future states as inferred posterior probability distributions
will allow to identify and classify anomalies (e.g. hardware failures,
software errors, communication, malicious activity, cyberattacks, e.t.c.) on a
higher level of model abstraction — the latent variables of the model, and
achieve finer and more robust anomaly detection and characterization.

>>>>

Modern information systems and operation environments are
commonly monitored through collection and streaming of
multivariate time series. The monitoring tasks comprise both
forecasting, for planning of resource allocation and decision
making, and novelty detection and characterization, for ensuring
faultless functioning and early mitigation of failures and
threats.  Advances in time series forecasting
due to adoption of multilayer recurrent neural network
architectures made it possible to forecast in high-dimensional
time series, and identify and classify novelties (anomalies)
early, based on subtle changes in the trends.  However,
mainstream approaches to multi-variate time series modelling
do not handle well cases when uncertainty is involved, either in
the input, when some of the observations are missing, or in the
output when the distribution of future observations, rather than
their point values, is predicted. For forecast uncertainty
modelling, stochastic latent variable
variants of high-dimensional time series models where
introduced, but so far have had to rely on sampling to account
for uncertainty, limiting the performance of data handling.
Imputation schemes were proposed for dealing with 
missing data, however, they do not generally give a satisfactory
solution in presence of transient unavailability of some of
the data sources (e.g. when a sensor stops working, or a
transport channel malfunctions), which is  a common case with
monitoring of complex systems.

A systematic and theoretically founded approach to handling
both input and output uncertainty would thus constitute a
significant and welcome contribution to the theory and practice
of monitoring of multivariate time series. It would also be
highly desirable for such approach to facilitate efficient
offline (learning) and online (inference) computations. In this
ongoing research, we propose a deep learning architecture which
uses a simple but powerful extension of traditional recurrent
neural network (RNN) architecture which allows both 
\begin{itemize} 
\item to handle missing inputs
in some or all of the components in a multivariate time series,
\item and to accomplish multi-step probabilistic forecasting
\end{itemize}
in high-dimensional time series, paving a path to better
decision making and finer and more robust anomaly detection and
characterization.  We evaluate the architecture on a real-world
data set of multivariate time series collected from a 
network for cloud computing, and empirically demonstrate
advantage of the proposed architecture over commonly used
approaches.
